---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# regutools

<!-- badges: start -->
[![Travis build status](https://travis-ci.org/ComunidadBioInfo/regutools.svg?branch=master)](https://travis-ci.org/ComunidadBioInfo/regutools)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
[![Codecov test coverage](https://codecov.io/gh/ComunidadBioInfo/regutools/branch/master/graphs/badge.svg)](https://codecov.io/gh/ComunidadBioInfo/regutools?branch=master)
<!-- badges: end -->

The goal of `regutools` is to provide an R interface for extracting and processing data from [RegulonDB](http://regulondb.ccg.unam.mx/).

## Installation

You can install the released version of `regutools` from [Bioconductor](http://bioconductor.org/) with:

```{r 'install release', eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("regutools")

## Check that you have a valid Bioconductor installation
BiocManager::valid()
```

And the development version from [GitHub](https://github.com/) with:

```{r 'install devel', eval = FALSE}
BiocManager::install("ComunidadBioinfo/regutools")
```

## Example

This is a basic example which shows you how to use `regutools`. For more details, please check the vignette.

```{r example}
library('regutools')
## basic example code

## Connect to the RegulonDB database if necessary
if(!exists('regulondb_conn')) regulondb_conn <- connect_database()

## Build a regulondb object
e_coli_regulondb <-
    regulondb(
        database_conn = regulondb_conn,
        organism = "E.coli",
        database_version = "1",
        genome_version = "1"
    )

## Get the araC regulators
araC_regulation <-
    get_gene_regulators(
        e_coli_regulondb,
        genes = c("araC"),
        format = "multirow",
        output.type = "TF"
    )

## Summarize the araC regulation
get_regulatory_summary(e_coli_regulondb, araC_regulation)
```

# Citation

Below is the citation output from using `citation('regutools')` in R. Please 
run this yourself to check for any updates on how to cite __regutools__.

```{r 'citation'}
citation('regutools')
```

# Testing

Testing on Bioc-devel is feasible thanks to [R Travis](http://docs.travis-ci.com/user/languages/r/).
